\section{Introduction}

In recent years, reinforcement learning (RL) has become a powerful paradigm for developing artificial intelligence systems capable of learning complex behaviors from experience. Unlike supervised learning approaches that require labeled data, RL enables agents to learn through trial and error by interacting with an environment and receiving feedback in the form of rewards \cite{sutton2018reinforcement}. This approach has led to remarkable achievements in various domains, including game playing \cite{mnih2015human, silver2017mastering}, robotics \cite{kober2013reinforcement}, and control systems \cite{lillicrap2015continuous}.

Game environments provide ideal testbeds for RL research due to their well-defined rules, clear objectives, and controllable complexity. They offer a balance between simplified abstractions and the complex dynamics found in real-world scenarios. Among these, 2D action games present a particularly interesting challenge, as they require agents to develop skills in spatial navigation, target acquisition, obstacle avoidance, and resource management—all while operating in continuous state spaces with partial observability.

\subsection{Project Context}

This project, AgentArena, aims to create a 2D shooting game environment specifically designed to serve as a testbed for different agent implementations, with a focus on reinforcement learning techniques. As a student of Applied Mathematics with interests in robotics and machine learning, this project represents an opportunity to apply theoretical knowledge to a practical implementation of RL in a dynamic, interactive environment.

The game provides a complete framework for developing, training, and evaluating agents, from simple rule-based implementations to sophisticated learning-based approaches. By creating agents capable of mastering this environment through reinforcement learning, we can gain insights into how AI systems develop strategic behaviors, spatial reasoning, and adaptive tactics.

\subsection{Research Question and Objectives}

The primary research question guiding this work is: \textit{How effectively can reinforcement learning agents develop strategies to succeed in a dynamic 2D shooting game environment with multiple objectives and adversaries?}

To address this question, the project has the following objectives:

\begin{itemize}
    \item Design and implement a flexible, modular game environment that serves as an effective testbed for AI agents
    \item Develop a reinforcement learning approach using Deep Q-Networks (DQN) capable of learning successful gameplay strategies
    \item Investigate the impact of different reward functions on agent behavior and performance
    \item Compare the performance of RL agents against random and rule-based baselines
    \item Analyze learned agent behaviors to identify emergent strategies and limitations
\end{itemize}

\subsection{Approach}

The AgentArena framework is built as a top-down 2D shooter where players navigate a grid-based arena, fire projectiles, and evade enemies. The system is designed with modularity as a core principle, separating the game logic from the agent implementations to facilitate experimentation with different AI approaches.

For the reinforcement learning implementation, we adopt the Deep Q-Network (DQN) architecture \cite{mnih2015human}, which combines Q-learning with deep neural networks to handle the high-dimensional state space of the game. The state representation includes the player's position and health, enemy positions and states, and bullet trajectories. The action space combines movement directions with shooting decisions, creating a complex decision space that the agent must learn to navigate effectively.

Multiple reward functions are implemented to explore how different incentive structures affect the agent's learning and resulting behavior, ranging from simple designs that reward hitting enemies and penalize taking damage to more sophisticated approaches that consider tactical positioning, resource management, and risk assessment.

\subsection{Overview of the Report}

This report presents the design, implementation, and evaluation of the AgentArena system and its reinforcement learning agents. Section 2 provides background on reinforcement learning and related work in game AI. Section 3 details the system architecture and core components. Section 4 describes the machine learning implementation, focusing on the DQN approach and state/action representations. Section 5 examines the different reward functions implemented and their theoretical implications. Section 6 presents experimental results and performance evaluations. Finally, Sections 7 and 8 discuss the findings, limitations, and future directions for this research.
